---
title: "Clustering: organize countries by socio-economic-health information"
author: "Advanced Modelling, Master in Computational Social Science"
date: 'UC3M, 2024'
output:
  html_document: 
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: no
    toc: no
    toc_depth: 1
  pdf_document:
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 1
editor_options:
  chunk_output_type: console
---


```{r global_options, include=T, echo = F}
knitr::opts_chunk$set(echo = T, warning=FALSE, message=FALSE)
```

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri(file.path("uc3m.jpg")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width="600",
               height="80")
```

# Motivation


Can we organize the countries of the world by variables we are interested in?

For instance, the World Bank organizes countries in 4 income groups using just the Gross National Income (GNI) per capita

But what if we are interested in health? And in much more variables?

In this case study, we will categorise countries using socio-economic and health indicators that determine the overall development of the world



```{r}
# delete everything
rm(list=ls()) 

library(tidyverse)
library(GGally) # ggplot2-based visualization of correlations
library(factoextra) # ggplot2-based visualization of pca
library(countrycode)
library(rworldmap)
library(mice)
library(plotly)
```

## The dataset 

The dataset is obtained through the notebook "WHO_preprocessing.Rmd", which creates a file named "WorldData2024.txt". There you can find explanation for the following World Health Organization (WHO) variables:

  - Life expectancy at birth (years)
  - Infant mortality rate (probability of dying between birth and age 1 per 1000 live births)
  - Population (in thousands) total
  - Ambient air pollution (Annual PM2.5 [ug/m3])
  - Estimates of rates of homicides per 100 000 population
  - Health expenditure as percentage of GDP (%)
  - Medical doctors (per 10 000 population)
  - Hospital beds (per 10 000 population)
  - Percentage of individuals using the Internet
  - Literacy rate (% of people ages 15 and above)
  - Expected Years Of School
  - Unemployment, total (% of total labor force)
  - GDP per capita (constant 2015 US$)
  

```{r}
WHO = read.table("./WorldData2024.txt", sep=",")  
head(WHO)
```

## Descriptive analysis

Some preliminar analysis

  
```{r}
# Enter your code here

```

Out of the 200 countries in the world, we have almost complete information (more than 10 variables) for around 130 countries: 

  - Diverse variables (socio-economic and health)

  - Missing values
  


The output of the tools depends directly on the selected variables for the input. This selection is subjective and depends on the application's objective

## Some pre-processing

What to do with missings?

```{r}
#Percentage of Na's?
sapply(WHO, function(x) sum(is.na(x))*100/nrow(WHO))

# Remove first countries with no population 
WHO = WHO %>% drop_na(Population)
```

Automatic imputation of NAs:

```{r}
m = 4 # number of multiple imputations
mice_mod <- mice(WHO, m=m, method='rf')
WHO <- complete(mice_mod, action=m)
```

Data frame:

```{r}
  load("./WHO.RData")

# Selecting only numerical variables from WHO since k-means only analyze the numberical variables.
X = WHO %>% dplyr::select(-Continent,-Region,-Country, -COUNTRY,-Population,-capital) 

# We need to apply the log to the GDP because it is highly assymetric
X$GDP = log(X$GDP)

# We just save the names and continents for the graphs
names=WHO$Country
continent = WHO$Continent
```  
  

# Clustering

Load important clustering libraries

```{r}
library(cluster)
install.packages("mclust")
library(mclust)
```  
  
Initial guess for the number of clusters: 5 (as the number of continents)  
  
```{r}
# Enter your code here
# centers : how many groups (5)
# nstart : how many times you want to repeat it (non random output than k = 5, 5 times repeat)
# fit : model name (it can be anything)
# in the output of the fit : 
# k-means clustering with 5 clusters of sizes --(67), --(24), --(50), --(52), --(23) (this is the continents that is clustered)
# Cluster means: LifeExpectancy 1 is the lifeexpectancy mean of continent of 67 (1st one) and so on.
fit = kmeans(X, centers = 5, nstart = 1000)

# Since with 5 clusters, overlapping happens a lot so we are gonna clustering 4 groups
fit = kmeans(X, centers = 4, nstart = 1000)

fit = kmeans(X, centers = 3, nstart = 1000)

View(WHO)
View(X)

```

Somehow an unbalanced classification

## Interpretation of centers:

```{r}
centers=fit$centers

barplot(centers[1,], las=2, col="darkblue")
barplot(centers[2,], las=2, col="darkblue")
barplot(centers[3,], las=2, col="darkblue")
barplot(centers[4,], las=2, col="darkblue")
barplot(centers[5,], las=2, col="darkblue")
```

One small group contains most dangerous countries

One big group seems to contain healthy countries

More insights?

## Clusplot 

Plot the countries in a 2D PCA graph, adding colors for the groups

```{r}
# clusplot
# put the countries according to their socio-economic characteristics
fviz_cluster(fit, data = X, geom = c("point"), ellipse.type = 'norm', pointsize = 1)+
  theme_minimal() + geom_text(label = names, hjust = 0, vjust = 0,size = 2, check_overlap = F) + scale_fill_brewer(palette = "Paired")
```

Remember the meaning of the first PCA: measure of global development (positive loads on good indicators and negative ones on the bad variables)


## Number of groups? 

We can get some hints using different methods: within cluster sums of squares, average silhouette and gap statistics.

```{r}
# y axis is the sum of the distance of components inside of the groups
# if there is the sudden change with spike or peaks then it gives the hint the optimal number of the clusters
fviz_nbclust(scale(X), kmeans, method = 'wss', k.max = 20, nstart = 1000)
# In this formual, the higher is the better
fviz_nbclust(scale(X), kmeans, method = 'silhouette', k.max = 20, nstart = 1000)
# nboot : number of Bootsrapping
# if the line is on specific number then it means it is the maximum number of clusters
fviz_nbclust(scale(X), kmeans, method = 'gap_stat', k.max = 10, nstart = 100, nboot = 100)
```

Any insight about the number of groups?

```{r}
# Enter your code here for your favourite number of groups
# In assignment, we can choose whatever number of groups but we need to demonstrate why we choose that number (clustering is very subjective)
fit.km = kmeans(X, centers = 3, nstart = 1000)
```

## Map the clustering in a map:
In this map we are gonna plot first putting the countries according to their geographical information, then we will cluster them
  
```{r}
# Select here your favorite clustering tool
map = data.frame(country=names, value=fit.km$cluster)
#map = data.frame(country=names, value=fit.kmeans$cluster)

#Convert the country code into iso3c using the function countrycode()
map$country = countrycode(map$country, 'country.name', 'iso3c')
#Create data object supporting the map
matched <- joinCountryData2Map(map, joinCode = "ISO3",
                               nameJoinColumn = "country")
#Draw the map
mapCountryData(matched,nameColumnToPlot="value",missingCountryCol = "white",
               borderCol = "#C7D9FF",
               catMethod = "pretty", colourPalette = "rainbow",
               mapTitle = c("Clusters"), lwd=1)
```

Insignts?

This is the result of the k-means which is the more or less best tool to see the clustering.
So the maps can show the socio-economic characteristics depending on the geographical location and
clustering plot shows the clustering depending on the socio-economic characteristics.

# Hierarchical clustering

Important to decide distance between observations and linkage to join groups

We need to decide first the distance and linkage

```{r}
# basic hierarchical clustering
d = dist(scale(X), method = "euclidean")
hc = hclust(d, method = "ward.D2")

# enhanced version
hc = eclust(X, "hclust", stand = T, hc_metric = "euclidean", hc_method = "ward.D2", graph = F)

```

## Visualization

Classical dendrogram:

```{r}
hc$labels <- names

fviz_dend(x = hc, 
          k=3,
          palette = "jco", 
          rect = TRUE, rect_fill = TRUE, cex=0.5,
          rect_border = "jco"          
)
```

Difficult to visualize the countries

Let's use a phylogenic tree:

```{r}
fviz_dend(x = hc,
          k = 3,
          color_labels_by_k = TRUE,
          cex = 0.8,
          type = "phylogenic",
          repel = TRUE)+  labs(title="Socio-economic-health tree clustering of the world") + theme(axis.text.x=element_blank(),axis.text.y=element_blank())
```

Insights?

Now in a geographical map

```{r}
groups.hc = cutree(hc, k = 8)

# Map our PCA index in a map:
map = data.frame(country=names, value=groups.hc)
#Convert the country code into iso3c using the function countrycode()
map$country = countrycode(map$country, 'country.name', 'iso3c')
#Create data object supporting the map
matched <- joinCountryData2Map(map, joinCode = "ISO3",
                               nameJoinColumn = "country")
#Draw the map
mapCountryData(matched,nameColumnToPlot="value",missingCountryCol = "white",
               borderCol = "#C7D9FF",
               catMethod = "pretty", colourPalette = "rainbow",
               mapTitle = c("Clusters"), lwd=1)

```


# Heatmaps

A heat map is a false color image (based on data frame X) with a dendrogram added to the left side and to the top

```{r}
heatmap(scale(X), scale = "none", labRow = names,
        distfun = function(x){dist(x, method = "euclidean")},
        hclustfun = function(x){hclust(x, method = "ward.D2")},
        cexRow = 0.7)
```

