---
title: "PCA: extracting socio-economic information in the world"
author: "Advanced Modelling, Master in Computational Social Science"
date: 'UC3M, 2024'
output:
  html_document: 
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: no
    toc: no
    toc_depth: 1
  pdf_document:
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 1
editor_options:
  chunk_output_type: console
---


```{r global_options, include=T, echo = F}
knitr::opts_chunk$set(echo = T, warning=FALSE, message=FALSE)
```

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri(file.path("uc3m.jpg")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width="600",
               height="80")
```

# Motivation

Can we organize the world's countries by variables we are interested in?

For instance, the World Bank organizes countries in 4 income groups using just the Gross National Income (GNI) per capita

But what if we are interested in health? And in much more variables?

In this case study, we will analyze countries using socio-economic and health indicators that determine the overall development of the world, and try to answer these questions

- How can we know who are the most developed countries?

- Is the socio-economic information related with the location one?


```{r}
# delete everything
rm(list=ls()) 

library(tidyverse)
library(GGally) # ggplot2-based visualization of correlations
library(factoextra) # ggplot2-based visualization of pca
library(countrycode)
library(rworldmap)
library(mice)
library(plotly)
```

## The dataset 

The dataset is obtained through the notebook "WHO_preprocessing.Rmd", which creates a file named "WorldData2024.txt". There you can find explanation for the following World Health Organization (WHO) variables:

  - Life expectancy at birth (years)
  - Infant mortality rate (probability of dying between birth and age 1 per 1000 live births)
  - Population (in thousands) total
  - Ambient air pollution (Annual PM2.5 [ug/m3])
  - Estimates of rates of homicides per 100 000 population
  - Health expenditure as percentage of GDP (%)
  - Medical doctors (per 10 000 population)
  - Hospital beds (per 10 000 population)
  - Percentage of individuals using the Internet
  - Literacy rate (% of people ages 15 and above)
  - Expected Years Of School
  - Unemployment, total (% of total labor force)
  - GDP per capita (constant 2015 US$)
  

```{r}
WHO = read.table("WorldData2024.txt", sep=",")  
```

## Descriptive analysis

Some preliminar analysis

  
```{r}
# Enter your code here

```

Out of the 200 countries in the world, we have almost complete information (more than 10 variables) for around 130 countries: 

  - Diverse variables (socio-economic and health)

  - Missing values
  


The output of the tools depends directly on the selected variables for the input. This selection is subjective and depends on the application's objective

## Some pre-processing

What to do with missings?

```{r}
#Percentage of Na's?
sapply(WHO, function(x) sum(is.na(x))*100/nrow(WHO))

# Remove first countries with no population 
WHO = WHO %>% drop_na(Population)
```

Automatic imputation of NAs:

```{r}
m = 4 # number of multiple imputations
mice_mod <- mice(WHO, m=m, method='rf')
WHO <- complete(mice_mod, action=m)
```

Data frame:

```{r}
X = WHO %>% dplyr::select(-Continent,-Region,-Country, -COUNTRY,-Population,-capital) 

# We need to apply the log to the GDP because it is highly assymetric
X$GDP = log(X$GDP)

# We just save the names and continents for the graphs
names=WHO$Country
continent = WHO$Continent
```  
  

## Basic Data Analysis


```{r}
# Add here your ideas for descriptive analysis

```

Which is the country with the lowest young population?

And the country with the highest senior population? Are they the same?
  
## Intelligent graphs with ggplot

For instance, this is 4D using a scatterplot (2D)

```{r}
ggplot(WHO, aes(x=MedicalDoctors, y=LifeExpectancy, group=Continent, size=Population, color=GDP)) + geom_point(alpha=0.9) + #geom_smooth(se=F, size=0.3) +
  facet_wrap(~ Continent) +
  scale_color_gradient(low="red", high="green") +
  theme_minimal()+ theme(legend.position="none") + 
  labs(title = "World countries: life expectancy vs MedicalDoctors", subtitle="(color denotes GDP, size denotes population)",caption="Source: World Health Organization",
       x = "MedicalDoctors", y = "Life expectancy at birth (in years)")
```

The same but adding interactive visualization with plotly:

```{r}
p=ggplot(WHO, aes(x=MedicalDoctors, y=LifeExpectancy, group=Continent, size=Population, color=GDP, text=names)) + geom_point(alpha=0.9) + #geom_smooth(se=F, size=0.3) +
  facet_wrap(~ Continent) +
  scale_color_gradient(low="red", high="green") +
  theme_minimal()+ theme(legend.position="none") + 
  labs(title = "World countries: life expectancy vs MedicalDoctors", subtitle="(color denotes GDP, size denotes population)",caption="Source: World Health Organization",
       x = "MedicalDoctors", y = "Life expectancy at birth (in years)")

ggplotly(p, tooltip=c("names"))
```

Is wealth a determinant of life expectancy?

```{r}
q = WHO %>% ggplot(aes(x=GDP, y=LifeExpectancy, group=Continent, size=Population, color=Region, text=names)) + geom_point(alpha=0.75) + scale_x_log10(labels=scales::comma) +
  facet_wrap(~ Continent) +
  theme_minimal()+ theme(legend.position="none") + 
  labs(title = "Life expectancy vs wealth by continent", subtitle="(size denotes population, color denotes region)",caption="Source: World Health Organization",
       x = "GDP", y = "Life Expectancy")
ggplotly(q, tooltip=c("names"))
```



## Motivation to PCA

What is the most developed country?

What is the country with the highest life expectancy? Is it the same?

What is the country with the lowest IMR?

Can we rank the countries considering all the variables at the same time?

## From dimension p to dimension 2

How to call PCA?

```{r}
# Enter your code here
pca =
  
```

With 2 components we explain 63% of variability

```{r}
library(factoextra)

fviz_screeplot(pca, addlabels = TRUE)
```

## First component?

Analyze first component


```{r}
# Enter your code here

```

It is a measure of global development (positive loads on good indicators and negative ones on the bad variables)

The most important indicators are: IMR, GDP, Internet, SchoolYears, LifeExpectancy, etc.

Let's plot squared loadings instead: hey are like percentages (numbers between 0 and 1), but we loose the sign

They are called contribution of variables to components

```{r}
fviz_contrib(pca, choice = "var", axes = 1)
```

The red dashed line on the graph above indicates the expected average contribution 

If the contribution of the variables were uniform, the expected value would be 1/length(variables) = 1/12 = 8%

Rank the countries by the first component (scores)

```{r}
names[order(pca$x[,1])][1:10]
names[order(pca$x[,1], decreasing=T)][1:10]
```

## Second component?

```{r}
barplot(pca$rotation[,2], las=2, col="darkblue")
```

Maybe we can get more insights by ranking the countries using this component:

```{r}
names[order(pca$x[,2])][1:5]
names[order(pca$x[,2], decreasing=T)][1:5]
```

Contribution of variables to second component

```{r}
fviz_contrib(pca, choice = "var", axes = 2)
```

## Plot the first two scores

```{r}
data.frame(z1=pca$x[,1],z2=pca$x[,2]) %>% 
  ggplot(aes(z1,z2,label=names,color=continent)) + geom_point(size=0) +
  labs(title="First two principal components (scores)", x="PC1", y="PC2") + #guides(color=guide_legend(title="HDI"))+
  theme_bw() +theme(legend.position="bottom") + geom_text(size=3, hjust=0.6, vjust=0, check_overlap = TRUE) 
# The two first PCs seem independent

```

Which are the regions with the most developed countries?

```{r}
data.frame(z1=-pca$x[,1],Region=WHO$Region) %>% group_by(Region) %>% summarise(mean=mean(z1), n=n()) %>% arrange(desc(mean))
```

## Conclusions?

Map:

```{r}
# Map our PCA index in a map:
map = data.frame(country=names, value=-pca$x[,1])
#Convert the country code into iso3c using the function countrycode()
map$country = countrycode(map$country, 'country.name', 'iso3c')
#Create data object supporting the map
matched <- joinCountryData2Map(map, joinCode = "ISO3",
                               nameJoinColumn = "country")
#Draw the map
mapCountryData(matched,nameColumnToPlot="value",missingCountryCol = "white",
               addLegend = FALSE, borderCol = "#C7D9FF",
               catMethod = "pretty", colourPalette = "terrain",
               mapTitle = c("PCA1 by Country"), lwd=1)


```



